---
title: "DLPFC data by 10x Visium (Maynard et al., 2021)"
author: "Zheng Li"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: 
  workflowr::wflow_htlm:
    toc: true
editor_options:
  chunk_output_type: console
---

# Introduction
Here, we apply BASS to analyze the human dorsolateral prefrontal cortex (DLPFC) 
data by 10x Visium from 
[Maynard et al., 2021](https://www.nature.com/articles/s41593-020-00787-0). 
DLPFC data contains expression values of 33,538 genes measured on two pairs of 
tissue sections from three independent neurotypical adult donors. Each pair 
consisted of two directly adjacent, 10  $\mu m$ serial tissue sections with the 
second pair located 300 $\mu m$ posterior to the first, resulting in a total of 
12 tissue sections. The original data can be downloaded from 
[here](https://github.com/LieberInstitute/spatialLIBD). We excluded spots that 
were not mapped to the tissue region in the histology image and retained a total 
of 33,538 genes measured on 4,226 (151507), 4,384 (151508), 4,789 (151509), 
4,634 (151510), 3,661 (151669), 3,498 (151670), 4,110 (151671), 4,015 (151672), 
3,639 (151673), 3,673 (151674), 3,592 (151675), and 3,460 (151676) spots along 
with their spatial locations for further analysis. The processed data can be 
found at the 
[data directory](https://github.com/zhengli09/BASS-Analysis/tree/master/data). 
We focused our analysis only on spatial domain detection because the clustering 
of spatial spots no longer has the cell type interpretation. For single-sample 
analysis, we analyzed each of the 12 tissue sections separately. For 
multi-sample analysis, we jointly analyzed four tissue sections from each adult 
donor because they contain similar tissue structures. 

# Load data and set hyper-parameters
```{r error=FALSE,message=FALSE}
library(BASS)
library(Matrix)
# cntm: a list of expression count matrices for 4 tissue sections
#       (151507-151510) from the same individual
# infom: a list of manually annotated labels of seven laminar 
#        clusters by the original study for 4 tissue sections
#        (151507-151510) from the same individual
# xym: a list of spatial coordinates for 4 tissue sections
#      (151507-151510) from the same individual 
load("data/spatialLIBD_p1.RData")

# hyper-parameters
# We set the number of cell types to a relatively large
# number (20) to capture the expression heterogeneity.
C <- 20
# number of spatial domains
R <- 7
```

# Run BASS
```{r warning=FALSE}
set.seed(0)
# Set up BASS object
BASS <- createBASSObject(cntm, xym, C = C, R = R,
  beta_method = "SW", init_method = "mclust", 
  nsample = 10000)

# Data pre-processing:
# 1.Library size normalization followed with a log2 transformation
# 2.Select top 3000 spatially expressed genes with SPARK-X
# 3.Dimension reduction with PCA
BASS <- BASS.preprocess(BASS, doLogNormalize = TRUE,
  geneSelect = "sparkx", nSE = 3000, doPCA = TRUE, 
  scaleFeature = FALSE, nPC = 20)

# Run BASS algorithm
BASS <- BASS.run(BASS)
```

```{r}
# post-process posterior samples:
# 1.Adjust for label switching with the ECR-1 algorithm
# 2.Summarize the posterior samples to obtain the spatial domain labels
BASS <- BASS.postprocess(BASS)
zlabels <- BASS@results$z # spatial domain labels
```

# Visualization
You can refer to 
[visualization](https://github.com/zhengli09/BASS-Analysis/tree/master/code/viz.R) 
for some useful plotting functions or you can write your own code for plotting.
```{r}
library(ggplot2)
source("code/viz.R")
```

```{r fig.height=4, fig.width=12}
zlabels <- lapply(zlabels, function(zlabels.l){
  zlabels.l <- factor(zlabels.l)
  levels(zlabels.l) <- c("L3", "L1", "L5", "L2", "L4", "WM", "L6")
  zlabels.l <- factor(zlabels.l, levels = c(
    "L1", "L2", "L3", "L4", "L5", "L6", "WM"
  ))
})
cols <- c("#264653", "#287271", "#2a9d8f", "#8ab17d", 
  "#e9c46a", "#f4a261", "#e76f51")
smps <- c("151507", "151508", "151509", "151510")
plotClustersFacet(xym, zlabels, smps, size = 1) +
  scale_color_manual("Domain", values = cols) +
  guides(color = guide_legend(nrow = 1, 
    override.aes = list(size = 2)))
```

